Supplemental Materials for the paper: \\ Pesticide exposure in small streams in Germany

### 2016-09-15 article.tex
second draft

\usepackage{todonotes}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% 150-200 words
Small streams are important refugia for biodiversity.
In agricultural areas, they may be at high risk of pesticide pollution. However, most related studies have been limited to a few streams on the regional level, hampering extrapolation to larger scales.
In Germany, pesticide monitoring is performed by the federal states as part of water quality surveillance. 
\end{abstract}

%% -------------------------------------------------------------------------
\section{Introduction}
\todo{Check all references to supplement}.
More than 50\% of the total land area in Germany are used by agriculture \citep{statistisches_bundesamt_bodenflache_2014}.
In the year 2014 more than 45,000 tonnes of 766 authorized pesticides were sold for application on this area \citep{bundesamt_fur_verbraucherschutz_und_lebensmittelsicherheit_absatz_2015}.
The applied pesticides may enter surface waters via spray-drift, edge-off-field run-off or drainage \citep{stehle_probabilistic_2013,schulz_comparison_2001,liess_determination_1999}.
Especially run-off after heavy precipitation events has been shown to be one of the major input routes for pesticides \citep{schulz_field_2004}.
Once entered the surface waters they may have adverse effects on biota and ecosystem functioning \citep{schafer_thresholds_2012}. 
Although, it is known that pesticides pollution and its effect increase with agricultural land-use \citep{schulz_field_2004}, studies investigating the shape of this relationships are missing.

\citet{malaj_organic_2014} analyzed data supplied to the European Union (EU) in the context of the Water Framework Directive (WFD) and showed that most European water bodies are at risk from pesticides.
\citet{stehle_pesticide_2015} compiled 1566 measured concentrations of 23 insecticides in the EU from scientific publications. 
They found that many of these measurements exceed regulatory acceptable concentrations (RAC).
Both studies indicate that pesticides might be a threat to biodiversity in the European union. 
However, theses studies reflect only a small amount of data and it is unclear how representative they are:
For Germany the study of \citet{malaj_organic_2014} lists only 175 sites and \citet{stehle_pesticide_2015} only 138 measurements. %175 estimated from digitized figure C.3 in Malaj 2014, Stehle: Table 2.
National monitoring programs are setup for the surveillance of water quality.
In Germany these are setup independently by the federal states in compliance with the WFD \citep{quevauviller_water_2008} and additional state specific needs. 
These programs may provide the most comprehensive data that is currently available for pesticide occurrences in German rivers.
However, currently there is no curated national-wide compilation of this monitoring data available.

Small water bodies (SWB) comprise a major fraction of streams \citep{nadeau_hydrological_2007}, accommodate a higher proportion of biodiversity compared to larger freshwater systems \citep{davies_comparison_2008, biggs_report_2014} and play an important role in recolonization of disturbed downstream reaches \citep{liess_analyzing_2005, orlinskiy_forested_2015}.
However, SWB might be also at high risk of pesticide contamination from adjacent agricultural areas and lower dilution potential \citep{schulz_field_2004,liess_determination_1999}.
It has been shown that SWB are more polluted than bigger streams \citep{stehle_pesticide_2015,schulz_field_2004}.
Especially, compounds from agricultural use show seasonal rain-event-driven peak concentrations in SWB \citep{wittmer_significance_2010}.
Despite their biological relevance and potential pesticide exposure only a small fraction of studies were conducted on pesticide pollution of SWB, with only few large scale studies \citep{lorenz_specifics_2016}. 
Moreover, it is currently unknown how precipitation might influence measured pesticide concentrations in national monitoring programs.

@@

A clear definition of SWB in terms of catchment or stream size is currently lacking \citep{lorenz_specifics_2016}. 
The WFD defines SWB with a catchment size between 10 and 100 km\textsuperscript{2}, without further consideration of streams \textless 10km\textsuperscript{2}. 
\citet{lorenz_specifics_2016} defines SWB with catchment size \textless 10km\textsuperscript{2} as SWB.
Because of data scarcity of streams \textless 10~km\textsuperscript{2} (Figure \ref{fig:fig3}) we define in this study all streams below 25~km\textsuperscript{2} as SWB. This catchment size corresponds to a stream width of approximately 2~meters (see Supplement Figure S2).

\subsection{Characterization of pesticide pollution}
We characterized pesticide pollution using regulatory acceptable concentrations (RAC) \citep{brock_linking_2010}.
RACs are derived during pesticide authorization and no unacceptable ecological effect are expected if the environmental concentration remains below this concentration.
The German Federal Environmental Agency provided RACs for the 105 compounds with highest detection rates (Supplement, Table S2). 
We expressed RACs as Risk Quotient (RQ):

@@

$\pi_i$ denotes the probability of a observation i being above LOQ and $f_{Gamma}$ denotes the gamma function and is used for values greater LOQ, where $\mu$ is the mean of the gamma function.
We used the precipitation at sampling date ($prec_0$) and the day before ($prec_{-1}$), as well as quarters of the year ($season_{Q1-Q4}$) as linear predictors for $\mu$ and $\pi$.
To account for temporal autocorrelation and differences between federal states we used the site nested within state as random intercepts:

\begin{align}
\begin{split}
\log(\mu_i) = prec_{0 ijk} + prec_{-1 ijk} + season_{ijk} + state_j + site_{jk}\\
logit(\pi_{ijk}) = prec_{0 ijk} + prec_{-1 ijk} + season_{ijk} + state_j + site_{jk}\\
state_{j} \sim N(0, \sigma^2_{state}) \\
site_{jk} \sim N(0, \sigma^2_{site})
\end{split}
\label{eqn:eqn4}
\end{align}

@@

In total 484 different compounds used as pesticides and their metabolites were measured at least once (Supplement, Table S2). 
Most of the compounds were herbicides (179), followed by insecticides (117) and fungicides (109).
Most samples were take in the months April till October, with less samples in the winter (see Supplemental Figure S2).
Only 5.5\% (160,800) of all measurements were detects above the limit of quantification (LOQ).
We found substantial differences in the spectra of analyzed compounds between federal states (Figure \ref{fig:fig2}).
Hierarchical clustering revealed three groups (see also Supplement Figure S2):

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figure2.pdf}
  \caption{Compound spectra of the different federal states. Left: Barcode plot - each vertical line is an analysed compound. Right: MDS ordination. 
  Colors according to three groups determined by hierarchical clustering (see Supplement Figure S2).}
  \label{fig:fig2}
\end{figure}

@@

\subsection{Effect of precipitation on pesticide exposure}
The spatio-temporal intersection revealed that 5\% of the samples were taken at or after days with rainfall events greater than 10mm / day (Supplement, Figure S3).
Precipitation at the day before ($precip_{-1}$) sampling increased $\mu$ and also $\pi$ for nearly all analysed compounds. 
The effects were less pronounced ($\pi$) or not directed ($\mu$) for precipitation at the day of sampling ($precip_{0}$)(Figure \ref{fig:fig5}, top).

Quarters 2-4 showed generally higher $\mu$ than in Q1. 
Exceptions were Diflufenican, that showed significantly lower RQ in Q2-Q4 compared to Q1.
Smaller negative effects were found for Clothianidin (Q2) and Isoproturon (Q3) (Figure \ref{fig:fig5}, bottom and Supplement Table S4).
We found similar effects on $\pi$.

@@

\subsection{Pesticides in small water bodies}
The dataset comprised 12,710 samples from 1295 small water bodies.
1173 samples (0.3\% of all and 5.6\% of samples with detects) RQ higher then 1 were observed.
Neonicotinoid Insecticides and Chlorpyrifos showed highest risk quotients (Figure \ref{fig:fig6}).
For Thiacloprid, Imidacloprid and Chlorpyrifos RAC was less than LOQ, therefore, all detections have a RQ \textgreater 1. 
The herbicides Nicosulfuron and Diflufenican, as well as the fungicide Dimoxystrobin showed also high exceedances of RQ (29.5/14.2/21.4 \% of all samples with detects).
Highest RQ were observed for Chlorpyrifos (max(RQ) = 244), Dimoxystrobin(max(RQ) = 117) and Isoproturon (max(RQ) = 80). 
Metabolites were most commonly detected in samples (Metazachlor sulfonix acid was in 82\% of all samples were it was analysed) (Supplement Figure S8)
Glyphosate was the compound with highest detection rates, followed by Boscalid and Isoproturon. 
However, only the latter showed RQ exceedances (Figure \ref{fig:fig6}).
In 44.8\% of samples more then one compound was quantified, with a maximum of 54 different compounds in one samples (supplemental figure S7).

@@

\todo{unfertig}

\subsection{Thresholds for agricultural land use and catchment size}
% agriculture
We found a strong influence of agriculture on the pollution of streams.
If there is more the 25\% agriculture within a catchment pesticides it is likely that RAC will be exceeded with an increase in fully agricultural catchments (above 75 \% agriculture).
To our knowledge this is the first study investigating such thresholds of pesticide exposure.
Thresholds for agricultural land use have only been investigated in the past for biological communities.
\citet{feld_response_2013} found change points of biological community metric for agricultural land use at 40\% in low lands. 
Similarly, \citet{waite_agricultural_2014} found a threshold for aquatic diatoms at 40\%.
Our results coincide with these thresholds and suggest that pesticides might contribute to the observed biological effects.

@@

\subsection{Effect of precipitation on pesticide exposure}
Our results revealed that pesticide sampling for chemical monitoring in Germany is mainly performed when no precipitation occurs. 
However, we found higher pesticide exposure if samples where taken at the after heavy rainfall events. 
Interestingly, samples take at they day of a heavy-rainfall-event did not show such a strong relationship.
This could be explained by a sampling just before the actual rainfall event.
Pesticide concentration in agricultural SWB show short term peak concentrations \citep{wittmer_significance_2010}.
Overall, our results indicate that current chemical monitoring strongly underestimates pesticide exposure.

\todo{add alternatives - passive sampler, automatic samplers...}

We found pesticide specific patterns in seasonality of the studied compounds.
Most pesticides showed an increase from April - September, which coincides with their main application season.
Hence, we also found the hightest RQ for the herbicide Diflufenican during the winter/spring season, which is application period for which it is registered \todo{cite BVL}.

\subsection{Pesticides in small water bodies}
Our results suggest that SWB are frequently exposed to biologically relevant pesticide concentrations.
\citet{stehle_pesticide_2015} found the highest percentage of RAC exceedances for organophosphate insecticides. 
By contrast, our results revealed that neonicotinoid insecticides show high exceedances, followed by the organophosphate chlorpyrifos. 
This difference can be attributed to the low sample size for neonicotinoid insecticides in their study (n = 33) compared to the dataset presented here (between 1290 and 2113 samples in SWB, Figure \ref{fig:fig6}). 
Our results shows that this particular class of insecticides may currently pose a high risk to freshwater ecosystems.
Compared to \citet{stehle_pesticide_2015} we found much lower rates RAC exceedances (0.3\% of 372,304 samples with RAC vs 44\% of 1,566 samples). 
This can be attributed to different aims of the data sources: scientific research aims at finding pollutants, whereas monitoring aims mainly at surveillance of water quality, also during periods with lower pesticide usage and at natural sites. 
Contrary, \citet{knauer_pesticides_2016} found exceedances from monitoring data mainly for herbicides and fungicides and only one insecticide Chlorpyrifos.
This might reflect differences in pesticide use between countries and defined RACs.

From the definition of RAC it follows that if the concentration of a compound exceeds its RAC ecological effects are expected.
Accordingly, \citet{stehle_agricultural_2015} found that biological diversity is significantly reduced at RQ of 1.12.
We found RQ values greater than 1.12 at 25\% of all SWB sites. 
Consequently, we conclude that pesticides from agricultural land use are a major treat to small water bodies, the biodiversity they host and services they provide. 
Additional treat arises when taking into account that most pesticides do not occur individually but in mixtures \cite{schreiner_pesticide_2016} and may co-occur with other stressors \citep{schafer_contribution_2016}.

% Approval / Risk Assessment
Monitoring data, despite the outlined limitations, provides an opportunity to study environmental occurrences of pesticides after approval.
The high exceedances of RAC indicate that the approval process for pesticides must be checked and refined.

>>>>

% \usepackage{todonotes}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
% 150-200 words
Small streams are important refugia for biodiversity.
In agricultural areas, they may be at high risk from pesticides that enter streams during rainfall. 
However, most related studies have been limited to a few streams on the regional level, hampering extrapolation to larger scales.
In Germany, pesticide monitoring is performed by the federal states as part of water quality surveillance.

We compiled monitoring data focusing on small streams, resulting in a data set of 2,918,604 measurements from 42,236 samples in 3,049 sampling sites covering the years 2005-2014.
A total of 484 different compounds that can be classified as pesticides were measured.
We studied the relationships between pesticide exposure and agricultural land use, catchment size, as well as precipitation and seasonal dynamics.

We found a threshold of 25\% agricultural land use, above pesticide exposure becomes apparent. 
Precipitation increased pesticide concentrations and these showed pesticide specific temporal patterns.
Especially neonicotinoid insecticides and Chlorpyrifos showed exceedances of regulatory acceptable concentrations (RACs) in small water bodies.

We conclude that pesticides from agricultural land use are a major treat to small water bodies, the biodiversity they host and services they provide.

% Total words: 
\end{abstract}

%% -------------------------------------------------------------------------
\section{Introduction}
More than 50\% of the total land area in Germany are used by agriculture \citep{statistisches_bundesamt_bodenflache_2014}.
In the year 2014 more than 45,000 tonnes of 766 authorized pesticides were sold for application on this area \citep{bundesamt_fur_verbraucherschutz_und_lebensmittelsicherheit_absatz_2015}.
The applied pesticides may enter surface waters via spray-drift, edge-off-field run-off or drainage \citep{stehle_probabilistic_2013,schulz_comparison_2001,liess_determination_1999}.
Especially run-off after heavy precipitation events has been shown to be one of the major input routes for pesticides \citep{schulz_field_2004}.
Once entered the surface waters they may have adverse effects on biota and ecosystem functioning \citep{schafer_thresholds_2012}. 
Although, it is known that pesticides pollution and its effect increase with agricultural land-use \citep{schulz_field_2004}, studies investigating the shape of this relationships are missing.

\citet{malaj_organic_2014} analyzed data supplied to the European Union (EU) in the context of the Water Framework Directive (WFD) and showed that most European water bodies are at risk from pesticides.
\citet{stehle_pesticide_2015} compiled 1,566 measured concentrations of 23 insecticides in the EU from scientific publications. 
They found that many of these measurements exceed regulatory acceptable concentrations (RAC).
Both studies indicate that pesticides might be a threat to biodiversity in the European union. 
However, theses studies reflect only a small amount of data and it is unclear how representative they are:
For Germany the study of \citet{malaj_organic_2014} lists only 175 sites and \citet{stehle_pesticide_2015} only 138 measurements. %175 estimated from digitized figure C.3 in Malaj 2014, Stehle: Table 2.
National monitoring programs are setup for the surveillance of water quality.
In Germany these are setup independently by the federal states in compliance with the WFD \citep{quevauviller_water_2008} and additional state specific needs. 
These programs may provide the most comprehensive data that is currently available for pesticide occurrences in German rivers.
However, currently there is no curated national-wide compilation of this monitoring data available.

Small water bodies (SWB) comprise a major fraction of streams \citep{nadeau_hydrological_2007}, accommodate a higher proportion of biodiversity compared to larger freshwater systems \citep{davies_comparison_2008, biggs_report_2014} and play an important role in recolonization of disturbed downstream reaches \citep{liess_analyzing_2005, orlinskiy_forested_2015}.
However, SWB might be also at high risk of pesticide contamination from adjacent agricultural areas and lower dilution potential \citep{schulz_field_2004,liess_determination_1999}.
It has been shown that SWB are more polluted than bigger streams \citep{stehle_pesticide_2015,schulz_field_2004}.
Especially, compounds from agricultural use show seasonal rain-event-driven short term peak concentrations in SWB \citep{wittmer_significance_2010}.
Despite their biological relevance and potential pesticide exposure only a small fraction of studies were conducted on pesticide pollution of SWB, with only few large scale studies \citep{lorenz_specifics_2016}. 
Moreover, it is currently unknown how precipitation might influence measured pesticide concentrations in national monitoring programs.

@@

A clear definition of SWB in terms of catchment or stream size is currently lacking \citep{lorenz_specifics_2016}. 
The WFD defines SWB with a catchment size between 10 and 100 km\textsuperscript{2}, without further categorisation of streams \textless 10km\textsuperscript{2}. 
\citet{lorenz_specifics_2016} defines SWB with catchment size \textless 10km\textsuperscript{2} as SWB.
Because of data scarcity of streams \textless 10~km\textsuperscript{2} (Figure \ref{fig:fig3}) we define in this study all streams below 25~km\textsuperscript{2} as SWB. This catchment size corresponds to a stream width of approximately 2~meters (see Supplement, Figure S2).

\subsection{Characterization of pesticide pollution}
We characterized pesticide pollution using regulatory acceptable concentrations (RAC) \citep{brock_linking_2010}.
RACs are derived during pesticide authorization as part of the ecological risk assessment.
No unacceptable ecological effect are expected if the environmental concentration remains below this concentration.
The German Federal Environmental Agency provided RACs for the 105 compounds with highest detection rates (Supplement, Table S2). 
We expressed RACs as Risk Quotient (RQ):

@@

$\pi_i$ denotes the probability of a observation i being above LOQ and $f_{Gamma}$ denotes the gamma function and is used for values greater LOQ, where $\mu$ is the mean of the gamma function.
We used the log(x+0.05) transformed precipitation at sampling date ($log~prec_0$) and the day before ($log~prec_{-1}$), as well as quarters of the year ($season_{Q1-Q4}$) as linear predictors for $\mu$ and $\pi$.
To account for temporal autocorrelation and differences between federal $state$ we used the $site$ nested within $state$ as random intercepts.

\begin{align}
\begin{split}
\log(\mu_{ijk}) = log~prec_{0 ijk} + log~prec_{-1 ijk} + season_{ijk} + state_j + site_{jk}\\
logit(\pi_{ijk}) = log~prec_{0 ijk} + log~prec_{-1 ijk} + season_{ijk} + state_j + site_{jk}\\
state_{j} \sim N(0, \sigma^2_{state}) \\
site_{jk} \sim N(0, \sigma^2_{site})
\end{split}
\label{eqn:eqn4}
\end{align}

@@

In total 484 different compounds used as pesticides and their metabolites were measured at least once (Supplement, Table S2). 
Most of the compounds were herbicides (179), followed by insecticides (117) and fungicides (109).
Most samples were take in the months April till October, with less samples in the winter (see Supplement, Figure S3).
Only 5.5\% (160,800) of all measurements were detects above the limit of quantification (LOQ).
We found substantial differences in the spectra of analyzed compounds between federal states (Figure \ref{fig:fig2}).
Hierarchical clustering revealed three groups (see also Supplemen,t Figure S4):

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{figure2.pdf}
  \caption{Compound spectra of the different federal states. Left: Barcode plot - each vertical line is an analysed compound. Right: MDS ordination. 
  Colors according to three groups determined by hierarchical clustering (see Supplement, Figure S4).}
  \label{fig:fig2}
\end{figure}

@@

\subsection{Effect of precipitation on pesticide exposure}
The spatio-temporal intersection revealed that 5\% of the samples were taken at or after days with rainfall events greater than 10mm / day (Supplement, Figure S6).
Precipitation at the day before ($log~precip_{-1}$) sampling increased $\mu$ and also $\pi$ for nearly all analysed compounds. 
The effects were less pronounced ($\pi$) or not clearly directed ($\mu$) for precipitation at the day of sampling ($log~precip_{0}$)(Figure \ref{fig:fig5}, top).

Q2, Q3 and Q4 showed generally higher $\mu$ than in Q1, with smallest differences for Q4.
An exception was the herbicide Diflufenican, that showed througout all quarters lower RQ in Q2-Q4 compared to Q1 (Figure \ref{fig:fig5}, bottom and Supplement Table S4).
We found similar effects of season on $\pi$, but with more compounds showing a reduction in $\pi$.

@@

\subsection{Pesticides in small water bodies}
The dataset comprised 12,710 samples from 1,295 small water bodies.
In 1173 samples (0.3\% of all and 5.6\% of samples with detects) RQ higher then 1 were observed.
Neonicotinoid Insecticides and Chlorpyrifos showed highest risk quotients (Figure \ref{fig:fig6}).
For Thiacloprid, Imidacloprid and Chlorpyrifos RAC was less than LOQ, therefore, all detections have a RQ~\textgreater~1. 
The herbicides Nicosulfuron and Diflufenican, as well as the fungicide Dimoxystrobin also showed high exceedances of RQ (29.5, 14.2 and 21.4 \% of all samples with detects).
Highest RQ were observed for Chlorpyrifos (max(RQ) = 244), Dimoxystrobin(max(RQ) = 117) and Isoproturon (max(RQ) = 80). 
Metabolites were most commonly detected in samples (Metazachlor sulfonic acid was in 82\% of all samples were it was analysed) (Supplement Figure S8).
Glyphosate was the compound with highest detection rates, followed by Boscalid and Isoproturon. 
However, only the latter showed RQ exceedances (Figure \ref{fig:fig6}).
In 44.8\% of samples more then one compound was quantified, with a maximum of 54 different compounds in one samples (Supplement, Figure S7).

@@

Given their occurrence in the landscape \citep{nadeau_hydrological_2007} SWB are underrepresented in the current monitoring (Figure \ref{fig:fig3}). 
Especially, streams below 10~km\textsuperscript{2} are missing, which could be attributed to the missing categorisation in the WFD.

\subsection{Thresholds for agricultural land use and catchment size}
% agriculture
We found a strong influence of agriculture on the pollution of streams.
If there is more the 25\% agriculture within a catchment pesticides it is likely that RAC will be exceeded with an increase in fully agricultural catchments (above 75 \% agriculture).
To our knowledge this is the first study investigating such thresholds of pesticide exposure.
Thresholds for agricultural land use have only been investigated in the past for biological communities.
\citet{feld_response_2013} found change points of biological community metric for agricultural land use at 40\% in low lands. 
Similarly, \citet{waite_agricultural_2014} found a threshold for aquatic diatoms at 40\%.
Our results coincide with these thresholds and suggest that pesticides might contribute to the observed biological changes.

@@

\subsection{Effect of precipitation on pesticide exposure}
Our results revealed that pesticide sampling for chemical monitoring in Germany is mainly performed when no precipitation occurs. 
Nevertheless, we found higher pesticide exposure if samples where taken at the day after heavy rainfall events. 
Interestingly, samples take at they day of the rainfall-event did not show such a strong relationship.
This could be explained by a sampling just before the actual rainfall event.
Pesticide concentration in agricultural SWB show short term peak concentrations \citep{wittmer_significance_2010}, therefore, a sampling at the day after the rain-fall-event might miss peak concentrations.
Overall, our results indicate that current chemical monitoring strongly underestimates pesticide exposure. 
Automatic event-drive samplers \citep{stehle_probabilistic_2013} and passive samplers \citep{fernandez_calibration_2014,moschet_evaluation_2015} may help overcome these shortcomings and provide a better representation, especially for SWB \citep{lorenz_specifics_2016}.

We found pesticide specific patterns in the seasonality of exposure.
Most pesticides showed an increase from April - September, which coincides with their main application season.
Hence, we also found the hightest RQ for the herbicide Diflufenican during the winter/spring season, which is application period for which it is registered in Germany \citep{bvl_online_2016}. 
Our study suggest that pesticide exposure shows compound specific spatio-temporal dynamics.
Currently, little is known on these and further research on those might provide useful information for future ecological risk assessment.

\subsection{Pesticides in small water bodies}
Our results suggest that SWB are frequently exposed to biologically relevant pesticide concentrations.
\citet{stehle_pesticide_2015} found the highest percentage of RAC exceedances for organophosphate insecticides. 
By contrast, our results revealed that neonicotinoid insecticides show high exceedances, followed by the organophosphate chlorpyrifos. 
This difference can be attributed to the low sample size for neonicotinoid insecticides in their study (n = 33) compared to the dataset presented here (between 1,290 and 2,113 samples in SWB, Figure \ref{fig:fig6}). 
Our results shows that this particular class of insecticides may currently pose a high risk to freshwater ecosystems.
Compared to \citet{stehle_pesticide_2015} we found much lower rates RAC exceedances (0.3\% of 372,304 samples with RAC vs 44\% of 1,566 samples). 
This can be attributed to different aims of the data sources: scientific research aims at finding pollutants, whereas monitoring aims mainly at surveillance of water quality, also during periods with lower pesticide usage and at natural sites. 
Contrary, \citet{knauer_pesticides_2016} found exceedances from monitoring data mainly for herbicides and fungicides and only one insecticide Chlorpyrifos.
This might reflect differences in pesticide use between countries and defined RACs.

From the definition of RAC it follows that if the concentration of a compound exceeds its RAC ecological effects are expected.
Accordingly, \citet{stehle_agricultural_2015} found that biological diversity is significantly reduced at a RQ of 1.12.
We found RQ values greater than 1.12 at 25\% of all SWB sites. 
Consequently, we conclude that pesticides from agricultural land use are a major treat to small water bodies, the biodiversity they host and services they provide. 
Additional treat arises when taking into account that most pesticides do not occur individually but in mixtures \cite{schreiner_pesticide_2016} and may co-occur with other stressors \citep{schafer_contribution_2016}.

% Approval / Risk Assessment
Monitoring data, despite the outlined limitations, provides an opportunity to study large scale environmental occurrences of pesticides.
Nevertheless, such nationwide compilations, may not only be used for governmental surveillance, but also to answer other questions, like validation of exposure modeling \cite{knabel_fungicide_2014}, retrospective evaluation of regulatory risk assessment \citep{knauer_pesticides_2016,stehle_pesticide_2015}or occurrences of pesticide mixtures \cite{schreiner_pesticide_2016}. 
The high exceedances of RAC indicate that the approval process for pesticides must be checked and refined.

### 2016-09-15 supplement/Supplemental_Materials.tex
second draft

%% -------------------------------------------------------------------------
\title{Supplemental Materials for the paper: \\Pesticides pollution of small streams in
Germany}
\author{Eduard Szöcs, Marvin Brinke, Bilgin Karaoglan, Ralf B. Schäfer}

@@

\begin{enumerate}
	\item Structure: Structure has been adjusted to the database structure.
	\item Coordinates: Coordinates have been transformed to a common Coordinate Reference System (DHDN / 3-Grad Gauss-Krüger Zone 3 (EPSG:31467) and duplicates merged.
	\item Chemicals: Chemical names and identifiers have been unified using the webchem package \citep{szocs_webchem:_2016}.
	\item  Identifiers: Unique identifiers have been assigned.
	\item Units: All concentrations have been converted to $\mu g/L$. Values below limit of quantification have be set to zero.
	\item Other meta-data: meta-data has been standardised.
	\item Temporal resolution: The temporal resolution of the database is 1 day. Date below this resolution has been aggregated by maximum.
	\item Validity Checks: Simple rules for validity checks have been implemented (e.g. no negative concentrations).
\end{enumerate}

@@

%% -------------------------------------------------------------------------
\chapter{Thresholds for agricultural land use and catchment size}
\begin{figure}[h]
	\centering
	\includegraphics[width = 0.8\textwidth]{ezgagrirac}
	\caption{Raw data used for the model in equation 2 and Figure 3 of the main article. Color codes the number of RAC exceedances (on a log-scale). Grey points denote sites without any exceedance.}
	\label{fig:ezgagrirac}
\end{figure}

%% -------------------------------------------------------------------------
\chapter{Effect of precipitation and season on RQ}
\begin{figure}[h]
	\centering
	\includegraphics[width = 0.8\textwidth]{precip}
	\caption{Distribution of precipitation at sampling occasions. top: at sampling date. bottom: at day before sampling.}
	\label{fig:precip}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width = 0.8\textwidth]{pdetects}
	\caption{Proportion of samples with detects in SWB. Only Compounds with more than 100 samples and 15\% of detects are show.}
	\label{fig:pdetects}
\end{figure}

>>>>

%% -------------------------------------------------------------------------
\title{Supplemental Materials for the paper: \\ Pesticide exposure in small streams in Germany}
\author{Eduard Szöcs, Marvin Brinke, Bilgin Karaoglan, Ralf B. Schäfer}

@@

\begin{enumerate}
	\item Structure: Structure has been adjusted to the database structure.
	\item Coordinates: Coordinates have been transformed to a common Coordinate Reference System (DHDN / 3-Grad Gauss-Krüger Zone 3 (EPSG:31467)) and duplicates merged.
	\item Chemicals: Chemical names and identifiers have been unified using the webchem package \citep{szocs_webchem:_2016}.
	\item  Identifiers: Unique identifiers have been assigned.
	\item Units: All concentrations have been converted to $\mu g/L$. Values below limit of quantification have be set to zero.
	\item Other meta-data: meta-data has been standardised.
	\item Temporal resolution: The temporal resolution of the database is 1 day. Date below this resolution has been aggregated by maximum.
	\item Validity Checks: Simple rules for validity checks have been implemented (e.g. no negative concentrations).
\end{enumerate}

@@

%% -------------------------------------------------------------------------
\chapter{Thresholds for agricultural land use and catchment size}
\begin{figure}[h]
	\centering
	\includegraphics[width = 0.8\textwidth]{ezgagrirac}
	\caption[Raw data used for the model in equation 2 and Figure 3 of the main article.]{Raw data used for the model in equation 2 and Figure 3 of the main article. Color codes the number of RAC exceedances (on a log-scale). Grey points denote sites without any exceedance.}
	\label{fig:ezgagrirac}
\end{figure}

%% -------------------------------------------------------------------------
\chapter{Effect of precipitation and season on RQ}
\begin{figure}[h]
	\centering
	\includegraphics[width = 0.8\textwidth]{precip}
	\caption[Distribution of precipitation at sampling occasions.]{Distribution of precipitation at sampling occasions. top: at sampling date. bottom: at day before sampling.}
	\label{fig:precip}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width = 0.8\textwidth]{pdetects}
	\caption[Proportion of samples with detects in SWB.]{Proportion of samples with detects in SWB. Only Compounds with more than 100 samples and 15\% of detects are show.}
	\label{fig:pdetects}
\end{figure}
